---
title: "ISYE 6501 HW 1"
output:
  pdf_document: default
  html_notebook: default
---

Last Modified 09/01/21

**Question 2.2**

Credit Card Data
```{r}
library(rmarkdown)

cc_data <- read.delim("~/Documents/R/data 2.2/credit_card_data.txt", #creating df
                      stringsAsFactors = FALSE, 
                      header=FALSE) 
data <- as.matrix(cc_data) #creating matrix
```

**Question 2.2.1**

Creating SVM model and calculating coefficients and intercept:
```{r}
library(kernlab)

set.seed(2) 

C = c(.001, 1, 100, 10000) #Values of C to be tested
accurracy = rep(0,4) #empty vector for accuracy values

# call ksvm in a loop to test several value of C.  Vanilladot is a simple linear kernel.  
for (i in C){
model <- ksvm(data[,1:10],
              data[,11],
              type='C-svc',
              kernel='vanilladot',
              C=i,
              scaled=TRUE)
pred <- predict(model,data[,1:10])
print(paste("Percent of model's prediction match actual value for C=", 
            i,
            "is", 
            sum(pred == data[,11]) / nrow(data)))
}
```
1 is selected as the best value for C with an accuracy of approximately 86%.

```{r}
#model with C=1
model <- ksvm(data[,1:10],
              data[,11],
              type='C-svc',
              kernel='vanilladot',
              C=1,scaled=TRUE)

# calculate a1…am
a <- colSums(model@xmatrix[[1]] * model@coef[[1]])
a

# calculate a0
a0 <- -model@b
a0
```
The kvsm classifier produces the following model:

$(.08148382 -0.0011026642 V1 - 0.0008980539 V2 - 0.0016074557 V3 + 0.0029041700 V4 + 1.0047363456 V5$ 
$- 0.0029852110 V6 - 0.0002035179 V7 - 0.0005504803 V8 - 0.0012519187 V9 + 0.1064404601 V10 + 0.08148382)1 = 0$

**Question 2.2.1 CONCLUSION** 

A ksvm classifier model trained on this data accurately predicts the value of the response variable for approximately 86% of the data points when we set a value of C = 1 and when C=100. 86% is the highest level of accuracy achieved in tests of different value of C. Lower values of C represent a wider margin of classification and higher values represent a more narrow margin. So it seems that 1 is a better choice than 100 since the margin for classifying points will be wider with the same classification accuracy.

**Question 2.2.3**

Creating a function that will iterate through all rows of the dataframe, for each, producing a k-Nearest-Neighbor model and returning it's accuracy using the Leave-one-out Cross Validation (LOOCV) method:

```{r}
library(kknn)

set.seed(2)

loocv <- function(x){ 
  predicted <- rep(0,(nrow(cc_data))) #establishing an empty vector
  for (i in 1:nrow(cc_data)){
  model_kknn <- kknn(V11~.,
                  train = cc_data[-i,],
                  test = cc_data[i,],
                  k = x,
                  distance = 2,
                  kernel = "optimal",
                  scale = TRUE)
  predicted[i] <- as.integer(fitted(model_kknn)+0.5)
}
  acc = sum(predicted == cc_data[,11]) / nrow(cc_data)
  return(acc)
}
```

Calling the function and applying it to the dataframe:
```{r}
accurracy=rep(0,25) #establishing an empty vector

for (i in 1:25){
  accurracy[i] = loocv(i)
}
```

Accuracy vs. K-Values:
```{r}
k_values <- c(1:25)

df <- data.frame(k_values, accurracy)

which.max(df[,2]) # find k-value with greatest accuracy value.

plot(df)
```
**Question 2.2.3 CONCLUSION** 

The LOOCV loop above creates a vector containing accuracy values for each value of k from 1-25.  The k-value with the highest accuracy is 12 as can be seen above.  This means that the k-Nearest-Neighbors model with the highest accuracy is one that classifies data points into clusters of 12 points nearest the centroid.
